---
title: "Week 4 Assessment Exercise 4"
author: "Mariam Serhan"
date: "2026-10-05"
output: 
  pdf_document:
    toc: true
fontsize: 11pt
citecolor: darkgreen
urlcolor: blue
bibliography: supportingfiles/references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Introduction to the NHANES Data Set 

The National Health and Nutrition Examination Survey (NHANES) is a program run by the CDC's National Center for Health Statistics (NCHS) and it purpose is to assess the health and nutrition of the American population. [@NHANES] Their data is released to the public and can be aquired using this link here: https://wwwn.cdc.gov/nchs/nhanes/ 

In this assignment, waves 2015, 2016 and 2017 will be used. 

# Importing Data and Loading Packages

```{r importdata}

library(pacman)

pacman::p_load(rio, here, dplyr, ggplot2, DT, tidyverse, formatR, gridExtra,
               pandoc, skimr, knitr, pandoc, flexdashboard, gt, flextable, 
               reactable, kableExtra, janitor, ggpubr)

cleaned_NHANES <- import(here("data", "cleaned_NHANES.csv"))
cleaned_NHANES_avrg_SBP <- import(here("data", "cleaned_NHANES_avrg_SBP.csv"))

```
# Creating Complex Tables

## Overall sample size of each wave

``` {r table_samplesize}

# sample size of each wave

table_size_wave <- table(cleaned_NHANES$wave_id)
table_size_wave

# format table
table_size_wave <- kable(
  table_size_wave, 
  format = "latex",
  booktabs = TRUE,
  col.names = c("Wave ID", "Count"),
  caption = "NHANES sample size of each wave"
  
)

# style table

table_size_wave <- table_size_wave %>%
	kable_styling(
		latex_options = "hold_position",
		full_width = FALSE 
)
table_size_wave

```


## Number and percentage of males

``` {r tables_males}

# sample size of each wave by gender

count_wave_gender <- table(cleaned_NHANES$wave_id, cleaned_NHANES$gender)
count_wave_gender

addmargins(count_wave_gender, margin = 2)

# row percentages

row_percent_gender <- prop.table(count_wave_gender, margin = 1) * 100
row_percent_gender

#combine and only keep male stats

male_percent_size <- as.data.frame(cbind(
  n = rowSums(count_wave_gender),
  Male = round(row_percent_gender[, 2], 1)
))
male_percent_size

# Format table

male_percent_size <- kable(
  male_percent_size, 
  format = "latex",
  booktabs = TRUE,
  col.names = c("Count", "Percent Male"),
  caption = "Percentage of male participants by wave year"
)

# style table

male_percent_size <- male_percent_size %>%
	kable_styling(
		latex_options = "hold_position",
		full_width = FALSE 
)
male_percent_size
```

## Number and percentage of each ethnicity for each wave

``` {r tables_ethnicity}

# sample size of each wave by ethncity 

count_wave_ethn <- table(cleaned_NHANES$wave_id, cleaned_NHANES$ethnicity_2)
count_wave_ethn

addmargins(count_wave_ethn, margin = 2)

# row percentages

row_percent_ethn <- prop.table(count_wave_ethn, margin = 1) * 100
row_percent_ethn

#combine 

ethn_percent_size <- as.data.frame(cbind(
  n = rowSums(count_wave_ethn),
  Asian = round(row_percent_ethn[, 1], 1),
  Black = round(row_percent_ethn[, 2], 1),
  Mexican = round(row_percent_ethn[, 3], 1),
  Other = round(row_percent_ethn[, 4], 1),
  "Other Hispanic" = round(row_percent_ethn[, 5], 1),
  White = round(row_percent_ethn[, 6], 1)
))

ethn_percent_size

# Format table

ethn_percent_size <- kable(
  ethn_percent_size, 
  format = "latex",
  booktabs  = TRUE,
  col.names = c(
    "Count", 
    "Asian", 
    "Black", 
    "Mexican", 
    "Other", 
    "Other Hispanic", 
    "White"),
  caption = "Count and percentage of each ethnicity for each wave"
)

ethn_percent_size <- ethn_percent_size %>%
	kable_styling(
		latex_options = "hold_position", 
		full_width = FALSE
	) %>%
  add_header_above(
		c(" " = 1, " " = 1, "Ethnicity" = 6)
	)

ethn_percent_size
```




## Summary statistics of average systolic blood pressure, by wave

``` {r tables_summary}

#generate summary

summary_stats <- cleaned_NHANES_avrg_SBP %>%
  group_by(wave_id) %>%
  summarise(
    min = round(min(SBP_avg, na.rm = TRUE), 1),
    max = round(max(SBP_avg, na.rm = TRUE), 1),
    median = round(median(SBP_avg, na.rm = TRUE), 1),
    mean = round(mean(SBP_avg, na.rm = TRUE), 1),
    sd = round(sd(SBP_avg, na.rm = TRUE), 1)
  )

summary_stats

# format Table

summary_stats <- kable(
  summary_stats, 
  format = "latex",
  booktabs  = TRUE,
  col.names = c(
    "wave ID", 
    "minimum", 
    "maximum", 
    "median", 
    "mean", 
    "SD"),
  caption = "Summary statistics of average systolic blood pressure, by wave"
)

summary_stats <- summary_stats %>%
	kable_styling(
		latex_options = "hold_position", 
		full_width = FALSE
	) %>%
  add_header_above(
		c(" " = 1, "Summary Statistic" = 5)
	)

summary_stats
```

# Bibliography